ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2024-12-02 paper/Manuscript-Data-Package.Rmd
Paper graphs and table update.

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes the open data product {ActiveCA}. {ActiveCA} is an `R` data package that contains analysis-ready data from the General Social Survey cicles regarding active travel in Canada. The package provides data set detailing active trip origins, destinations, and duration, covering a wide variety of locations, such as home, work or school, libraries, museums, restaurants, bars, sports centers, health clinics, place of worship, and others. The package also details the respondent's region characteristics, specifying wheter they live in a metropolitan area and their province of residency. {ActiveCA} also provides pre-estimated impedance functions for active travel for Canadian Metropolitan Area and Census Agglomerations, which can be used in accessibility analysis to calculate the cost of travel between different locations. The package will continue to expand with contributions from the authors and the broader community through requests in the future. {ActiveCA} is freely accessible for exploration and download from the associated Github repository, where the documentation and code involved in creating and manipulating data are detailed."

@@

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  comment = '', 
  out.width = "1\\linewidth")
```

@@

@@

This paper presents the open data product {ActiveCA}. Open data products
(ODPs) are the outcome of a transparent process that transforms raw data
(open or not) into analysis-ready data, in which all stages of
development follow open principles [@arribas-bel2021]. ODPs differ from
general open data due to their high utility, added value and open
availability.

{ActiveCA} is an `R` data package that provides analysis-ready data from the General Social Survey of Canada Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) [@statisticscanada2024] regarding active travel. These data can be used to study walking and cycling in Canada for various purposes over time, and to calculate impedance functions for accessibility analysis. The package provides pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas.

@@

The aim of this paper is to walk readers through the data sets and
invite others to experiment in its uses and applications. {ActiveCA} is
freely available on GitHub for all to install and freely use in the
spirit of open and reproducible research. In addition of enabling to obtain impedance functions, {ActiveCA} can be adopted in various applications that even go beyond the range of possibilities we have imagined. Not only the data, but also all the code documenting the processing methodology is available for consultation and
evaluation in its repository. This package contributes to reducing the
barrier to using the information contained in GSS surveys to provide
data-driven decisions in transportation analysis.

@@

Considering GSS Cycles analyzed, we identified `r nrow(gss_episodes)` episodes that recorded active travel episodes, with trip duration ranging from 0 to 900 minutes, to twelve different destinations. {ActiveCA} includes all these episodes ready for analysis. Table \ref{tab:table-01} presents descriptive statistics on walking and cycling trips between 1986 and 2015, including metrics such as the count
of recorded trips (count), and measures of trip duration in minutes:
maximum (max), mean, median, and minimum (min). The 1986 survey did not
include bicycle trips.

```{r table-01, echo=FALSE, cache=FALSE}
# Calculate the count of trips for each mode and year
episodes_counts <- gss_episodes %>%
  group_by(YEAR, MODE) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = YEAR, values_from = count)

# Calculate min, mean, and max statictis (duration)
stats_data <- gss_episodes %>%
  group_by(YEAR, MODE) %>%
  summarise(
    min = min(DURATION),
    mean = mean(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(min, mean, median, max), names_to = "Statistic", values_to = "Value") %>%
  pivot_wider(names_from = YEAR, values_from = Value)

# Combine counts and stats into one data frame
final_data <- bind_rows(
  episodes_counts %>% mutate(Statistic = "count"),
  stats_data
) %>%
  mutate(MODE = factor(MODE, levels = c("Walking", "Cycling"))) %>%
  arrange(MODE, Statistic) %>%
  mutate(across(where(is.numeric), ~ round(.x, 0)))

# Rename the 'MODE' column to 'Mode' before generating the table
final_data <- final_data %>%
  rename(Mode = MODE) %>%
  dplyr::select(Mode, Statistic, everything())

@@

wide_data <- wide_data %>%
  group_by(MODE) %>%
  mutate(MODE = if_else(row_number() == 2, as.character(MODE), "")) %>%
  ungroup() %>%
 dplyr::select(
    dest_label, MODE,
    min_1986, median_1986, max_1986, Percentage_1986,
    min_1992, median_1992, max_1992, Percentage_1992,
    min_1998, median_1998, max_1998, Percentage_1998
  )

# Latex table
styled_ch03_table_02 <- wide_data %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(wide_data))),
    caption = "\\label{tab:table-02}Comparison of travel statistics by mode and destination: 1986, 1992, 1998",
    col.names = c("Destination", "Mode*","Min*","Med*","Max*", "(%)*", "Min","Med","Max", "(%)", "Min","Med","Max", "(%)")
  ) %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 6) %>%
  # column_spec(1, width = "1cm") %>%  # Set width of 'Dest' column to 1.5cm
  # column_spec(2:14, width = "0.5cm") %>%  # Set width of remaining columns to 0.7cm
  add_header_above(c(" " = 2, "1986" = 4, "1992" = 4, "1998" = 4)) %>%
  row_spec(0, bold = TRUE, align = "c") %>%
  column_spec(6, border_right = TRUE) %>%
  column_spec(10, border_right = TRUE) %>% 
  footnote(general = c("* The symbols used in this table represent the following: `Min` denotes the minimum time to reach the destination; `Max` denotes the maximum time to reach the destination; `(%)` indicates a percentage of the total time to reach the destination; `Med` refers to the median time to reach the destination"), threeparttable = T)

@@

```{r table-03,echo=FALSE, cache=FALSE, warning=FALSE}
# Renaming 'dest_label' and 'orig_label' values to reduce table space
gss_episodes_renamed <- gss_episodes %>%
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))

# Selecting surveys after 2000 and calculating trip duration statistics 
episodes_stats_after <- gss_episodes_renamed %>%
  filter(YEAR > 2000) %>%
  group_by(YEAR, dest_label, MODE) %>%
  summarise(
    count = n(),
    min = min(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  )

@@

# Pivot data for the table layout
wide_data <- episodes_stats_after %>%
  pivot_wider(
    id_cols = c(dest_label, MODE),
    names_from = YEAR,
    values_from = c(min, median, max, Percentage)
  ) %>%
 dplyr::select(
    dest_label, MODE,
    min_2005, median_2005, max_2005, Percentage_2005,
    min_2010, median_2010, max_2010, Percentage_2010,
    min_2015, median_2015, max_2015, Percentage_2015
  ) %>% 
  arrange(MODE, dest_label, desc(Percentage_2015)) %>%
   group_by(MODE) %>%
   mutate(MODE = if_else(row_number() == 2, as.character(MODE), "")) %>%
   ungroup()

# Latex table
styled_ch03_table_03 <- wide_data %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(wide_data))),
    caption = "\\label{tab:table-03}Comparison of travel statistics by mode and destination: 2005, 2010, 2015",
    col.names = c("Destination", "Mode*","Min*","Med*","Max*", "(%)*", "Min","Med","Max", "(%)", "Min","Med","Max", "(%)")
  ) %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                font_size = 6) %>%
  column_spec(1, width = "2cm") %>%  # Set width of 'Dest' column to 1.5cm
  column_spec(2:14, width = "0.4cm") %>%  # Set width of remaining columns to 0.7cm
  add_header_above(c(" " = 2, "2005" = 4, "2010" = 4, "2015" = 4)) %>%
  row_spec(0, bold = TRUE, align = "c") %>%
  column_spec(6, border_right = TRUE) %>%
  column_spec(10, border_right = TRUE) %>% 
  footnote(general = c("* The symbols used in this table represent the following: `Min` denotes the minimum time to reach the destination; `Max` denotes the maximum time to reach the destination; `(%)` indicates a percentage of the total time to reach the destination; `Med` refers to the median time to reach the destination"), threeparttable = T)

@@

{ActiveCA} also enables visual analysis of active travel in Canada using traditional exploratory data analysis techniques. Figures \ref{fig:figure-01} and
\ref{fig:figure-02} show walking and cycling trips from 1992 and 2015
through heat maps. These maps use color gradients to represent the
percentage of trips between various origins and destinations, with
darker colors indicating higher percentages and lighter colors
representing less frequent routes. To avoid overwhelming the reader, we
omitted the heat maps for the other years analyzed.

@@

      # Heatmap for every combination of year and mode
      assign(paste0("heatmap_plot_", year, "_", mode),
               ggplot(heatmap_melted, aes(x = Destination, y = Origin, fill = value)) +
                  geom_tile(color = "white") +
                  scale_fill_gradientn(colors = RColorBrewer::brewer.pal(9, "YlOrRd"), 
                                       na.value = "gray95", 
                                       limits = c(0, ceiling(max(heatmap_melted$value, na.rm = TRUE))), 
                                       breaks =  c(0, ceiling(max(heatmap_melted$value, na.rm = TRUE))), 
                                       labels = c("0%", paste0(ceiling(max(heatmap_melted$value, na.rm = TRUE)), "%"))) +
                  geom_text(data = subset(heatmap_melted, !is.na(value) & value >= 0), 
                      aes(label = round(value)), 
                      size = 4, color = "black") +
                  labs(x = "Destination", y = "Origin", fill = "Percentage", caption = year) +
                  theme_minimal() +
                  theme(
                  axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
                  axis.text.y = element_text(size = 16, margin = margin(l = 15)),
                  axis.title = element_text(size = 16),
                  legend.title = element_text(size = 16, margin = margin(b = 15)), 
                  legend.text = element_text(size = 16),
                  plot.caption = element_text(hjust = 0.5, size = 20, face = "bold", margin = margin(t = 28.35, unit = "pt")))

@@

{ActiveCA} includes a total of
`r nrow(unique(gss_impedances[,c("YEAR","dest_label","f_name","MODE")]))`
pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas. Table \ref{tab:table-04} shows the best impedance functions for walking trips to `Work or school` across survey years, where all distributions, except for a gamma function in 1998, are log-normal. Figure
\ref{fig:figure-03} illustrates these fitted functions (red line)
alongside histograms of empirical travel times (grey bars). As for the
others pre-estimated functions, these examples allow to calculate gravity-based
accessibility measures for active travel across various destinations and temporal scales in Canadian urban areas.

```{r include = FALSE}
# Function to visualize impedance functions
visualize_impedance <- function(df_functions, travel_cost){

  x <- data.frame(t = seq(1, travel_cost, 1))

  for (year in unique(df_functions$YEAR)) {

    name_function <- df_functions[df_functions$YEAR == year, ]$f_name

    subset <- df_functions %>% 
      filter(YEAR == year)

    # new colunm with the year information in the name
    column_name <- paste0("year_", year)

    # lnorm distribution 
    if (name_function == "lnorm") {
      x <- x %>%
        mutate(!!sym(column_name) := dlnorm(t,
                                             meanlog = subset$est_1[1],
                                             sdlog = subset$est_2[1]))

    # Gamma distribution 
    } else if (name_function == "gamma") {

      x <- x %>%
        mutate(!!sym(column_name) := dgamma(t,
                                             shape = subset$est_1[1],
                                             rate = subset$est_2[1]))

    # Norm distribution

    }else if (name_function == "norm") {  
      x <- x %>%
        mutate(!!sym(column_name) := dnorm(t,
                                             mean = subset$est_1[1],
                                             sd = subset$est_2[1]))

    # Exponential distribution

    }else if (name_function == "exp") {  
      x <- x %>%
        mutate(!!sym(column_name) := dexp(t,
                                             exp = subset$est_1[1]))

    # Uniform distribution

    }else if (name_function == "unif") {  
      x <- x %>%
        mutate(!!sym(column_name) := dunif(t,
                                             min = subset$est_1[1],
                                             max = subset$est_2[1]))
    }
  }

  return(x)
}
```

```{r , include=FALSE}
# Setting the maximum time to be considered to plot the impedance graphs
max_time <- 100

# Loop to generate all impedance plots at once 
for(destino in unique(gss_impedances$orig_label)){

  for(mode in unique(gss_impedances[gss_impedances$orig_label == destino, ]$MODE)){

  selected_functions <-  gss_impedances %>% 
  filter(dest_label == destino, MODE == mode) %>%
  dplyr::select("YEAR", "f_name", "est_1", "est_2", "loglik", "aic", "bic") %>%
  mutate(across(c(est_1, est_2), ~ round(., digits = 2)))  %>%
  mutate(across(c(loglik, aic, bic), ~ round(., digits = 0)))  %>%
  unique()

  if(nrow(selected_functions) > 0){
    tld_theoretical <- visualize_impedance(df_functions = selected_functions, travel_cost = max_time)

  plots <- list()

# Loop through each unique year
  for(year in unique(gss_impedances$YEAR)){

    # Subset empirical and theoretical data for the current year
    empirical_subset <- gss_impedances %>% 
      filter(dest_label == destino, MODE == mode, YEAR == year, DURATION <= max_time)

    if(nrow(empirical_subset) > 1 ){

      print(paste0(mode," - ", destino," - ", year))

      theoretical_subset <- tld_theoretical %>%
      dplyr::select(t, matches(as.character(year)))

      # Create the ggplot
      plots[[as.character(year)]] <- ggplot(empirical_subset, aes(x = DURATION)) +
        geom_histogram(aes(y = ..density..), binwidth = 5, fill = "#dce1e3", color = "#adb1b3") +
        geom_line(data = theoretical_subset, aes(x = t, y = theoretical_subset[[2]]), color = "#b52640", size = 0.80) +
        labs(title = paste(year)) +
        theme_minimal() +
        scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) +
        theme(axis.title.x = element_blank(),
            axis.title.y = element_blank(), 
            plot.title = element_text(size = 10, face="bold"))

  }

# Use gridExtra to arrange the plots in accordance with the number of survey years (some destination exits only in specific surveys)

    # Case of only one year 
    if(length(plots) == 1){
      combined_plot <- grid.arrange(
      grobs = plots,
      ncol = 1,
      bottom = text_grob(paste0("Travel time to `", destino ,"` (in minutes) considering `", mode,"` as transportation mode.")),
      left = text_grob("Density", rot = 90)
    )
      ggsave(file = paste0(here::here(),"/paper/Manuscript-figures/impf_",destino,"_",mode,".jpg"), 
       plot = combined_plot,
       width = 9, 
       height = 4, 
       units = "in", 
       dpi = 300)

    # Destination present in 3 surveys  
    } else if(length(plots) == 3){
      combined_plot <- grid.arrange(
      grobs = plots,
      ncol = 3,
      bottom = text_grob(paste0("Travel time to `", destino ,"` (in minutes) considering `", mode,"` as transportation mode.")),
      left = text_grob("Density", rot = 90)
    )
      ggsave(file = paste0(here::here(),"/paper/Manuscript-figures/impf_",destino,"_",mode,".jpg"), 
       plot = combined_plot,
       width = 9, 
       height = 4, 
       units = "in", 
       dpi = 300)

>>>>

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes the open data product `ActiveCA`, with Canadian time use data about active travel. `ActiveCA` is an `R` data package that contains analysis-ready data from cycles  General Social Survey cicles regarding active travel in Canada. The package provides data set detailing active trip origins, destinations, and duration, covering a wide variety of locations, such as home, work or school, libraries, museums, restaurants, bars, sports centers, health clinics, place of worship, and others. The package also details the respondent's region characteristics, specifying wheter they live in a metropolitan area and their province of residency. `ActiveCA` also provides pre-estimated impedance functions for active travel for Canadian Metropolitan Area and Census Agglomerations, which can be used in accessibility analysis to calculate the cost of travel between different locations. The package will continue to expand with contributions from the authors and the broader community through requests in the future. `ActiveCA` is freely accessible for exploration and download from the associated Github repository, where the documentation and code involved in creating and manipulating data are detailed."

@@

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  comment = '', 
  out.width = "1\\linewidth")

opts <- options(knitr.kable.NA = "")
```

@@

```{r rename-labels, include = FALSE}
# Renaming 'dest_label' and 'orig_label' values to reduce table space
gss_episodes_renamed <- gss_episodes %>%
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))

gss_impedances_renamed <- gss_impedances %>%
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))
```

@@

This paper presents the open data product `ActiveCA`. Open data products
(ODPs) are the outcome of a transparent process that transforms raw data
(open or not) into analysis-ready data, in which all stages of
development follow open principles [@arribas-bel2021]. ODPs differ from
general open data due to their high utility, added value and open
availability.

`ActiveCA` is an `R` data package that provides analysis-ready data from the General Social Survey of Canada Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) [@statisticscanada2024] regarding active travel. These data can be used to study walking and cycling in Canada for various purposes over time, and to calculate impedance functions for accessibility analysis. The package provides pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas.

@@

The aim of this paper is to walk readers through the data sets and
invite others to experiment in its uses and applications. `ActiveCA` is
freely available on GitHub for all to install and freely use in the
spirit of open and reproducible research. In addition of enabling to obtain impedance functions, `ActiveCA` can be adopted in various applications that even go beyond the range of possibilities we have imagined. Not only the data, but also all the code documenting the processing methodology is available for consultation and
evaluation in its repository. This package contributes to reducing the
barrier to using the information contained in GSS surveys to provide
data-driven decisions in transportation analysis.

@@

Considering GSS Cycles analyzed, we identified `r nrow(gss_episodes)` episodes that recorded active travel episodes, with trip duration ranging from 0 to 900 minutes, to twelve different destinations. `ActiveCA` includes all these episodes ready for analysis. Table \ref{tab:table-01} presents descriptive statistics on walking and cycling trips between 1986 and 2015, including metrics such as the count
of recorded trips (count), and measures of trip duration in minutes:
maximum (max), mean, median, and minimum (min). The 1986 survey did not
include bicycle trips.

```{r table-01, echo=FALSE, cache=FALSE}
# Calculate the count of trips for each mode and year
episodes_counts <- gss_episodes %>%
  group_by(YEAR, MODE) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(names_from = YEAR, values_from = Count)

# Calculate min, mean, and max statictis (duration)
stats_data <- gss_episodes %>%
  group_by(YEAR, MODE) %>%
  summarise(
    min = min(DURATION),
    mean = mean(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    std = sd(DURATION),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(min, mean, median, max, std), names_to = "Statistic", values_to = "Value") %>%
  pivot_wider(names_from = YEAR, values_from = Value)

# Combine counts and stats into one data frame
final_data <- bind_rows(
  episodes_counts %>% mutate(Statistic = "Count"),
  stats_data
) %>%
  mutate(MODE = factor(MODE, levels = c("Walking", "Cycling"))) %>%
  arrange(MODE, Statistic) %>%
  mutate(across(where(is.numeric), ~ round(.x, 0)))

# Rename the 'MODE' column to 'Mode' before generating the table
final_data <- final_data%>% 
  arrange(MODE) %>%
  rename(Mode = MODE) %>%
  dplyr::select(Mode, Statistic, everything()) %>% 
   mutate(Statistic = case_when(
    Statistic == "max" ~ "Maximum",
    Statistic == "min" ~ "Minimum",
    Statistic == "mean" ~ "Mean",
    Statistic == "median" ~ "Median",
    Statistic == "std" ~ "Standard deviation",
    TRUE ~ Statistic  # Keep all other values as they are
  ))

@@

wide_data <- wide_data %>%
 dplyr::select(
    MODE, dest_label,
    min_1986, median_1986, max_1986, Percentage_1986,
    min_1992, median_1992, max_1992, Percentage_1992,
    min_1998, median_1998, max_1998, Percentage_1998
  ) %>% 
  arrange(MODE, dest_label)

# Latex table
styled_ch03_table_02 <- wide_data %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(wide_data))),
    caption = "\\label{tab:table-02}Comparison of travel statistics by mode and destination: 1986, 1992, 1998",
    col.names = c("Destination", "Mode*","Min*","Med*","Max*", "(%)*", "Min","Med","Max", "(%)", "Min","Med","Max", "(%)")
  ) %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 6) %>%
  # column_spec(1, width = "1cm") %>%  # Set width of 'Dest' column to 1.5cm
  # column_spec(2:14, width = "0.5cm") %>%  # Set width of remaining columns to 0.7cm
  add_header_above(c(" " = 2, "1986" = 4, "1992" = 4, "1998" = 4)) %>%
  row_spec(0, bold = TRUE, align = "c") %>%
  column_spec(6, border_right = TRUE) %>%
  column_spec(10, border_right = TRUE) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major") %>% 
  footnote(general = c("* The symbols used in this table represent the following: 'Min' denotes the minimum time to reach the destination; 'Max' denotes the maximum time to reach the destination; '(%)' indicates a percentage of the total time to reach the destination; 'Med' refers to the median time to reach the destination"), threeparttable = T)

@@

```{r table-03,echo=FALSE, cache=FALSE, warning=FALSE}
# Selecting surveys after 2000 and calculating trip duration statistics 
episodes_stats_after <- gss_episodes_renamed %>%
  filter(YEAR > 2000) %>%
  group_by(YEAR, dest_label, MODE) %>%
  summarise(
    count = n(),
    min = min(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  )

@@

# Pivot data for the table layout
wide_data <- episodes_stats_after %>%
  pivot_wider(
    id_cols = c(dest_label, MODE),
    names_from = YEAR,
    values_from = c(min, median, max, Percentage)
  ) %>%
 dplyr::select(
    MODE, dest_label,
    min_2005, median_2005, max_2005, Percentage_2005,
    min_2010, median_2010, max_2010, Percentage_2010,
    min_2015, median_2015, max_2015, Percentage_2015
  ) %>% 
  arrange(MODE, dest_label)

# Latex table
styled_ch03_table_03 <- wide_data %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(wide_data))),
    caption = "\\label{tab:table-03}Comparison of travel statistics by mode and destination: 2005, 2010, 2015",
    col.names = c("Destination", "Mode","Min*","Med*","Max*", "(%)*", "Min","Med","Max", "(%)", "Min","Med","Max", "(%)")
  ) %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                font_size = 6) %>%
  #column_spec(1, width = "2cm") %>%  # Set width of 'Dest' column to 1.5cm
  #column_spec(2:14, width = "0.4cm") %>%  # Set width of remaining columns to 0.7cm
  add_header_above(c(" " = 2, "2005" = 4, "2010" = 4, "2015" = 4)) %>%
  row_spec(0, bold = TRUE, align = "c") %>%
  column_spec(6, border_right = TRUE) %>%
  column_spec(10, border_right = TRUE) %>% 
  collapse_rows(columns = 1, valign = "top", latex_hline = "major") %>% 
  footnote(general = c("* The symbols used in this table represent the following: 'Min' denotes the minimum time to reach the destination; 'Max' denotes the maximum time to reach the destination; '(%)' indicates a percentage of the total time to reach the destination; 'Med' refers to the median time to reach the destination"), threeparttable = T)

@@

`ActiveCA` also enables visual analysis of active travel in Canada using traditional exploratory data analysis techniques. Figures \ref{fig:figure-01} and
\ref{fig:figure-02} show walking and cycling trips from 1992 and 2015
through heat maps. These maps use color gradients to represent the
percentage of trips between various origins and destinations, with
darker colors indicating higher percentages and lighter colors
representing less frequent routes. To avoid overwhelming the reader, we
omitted the heat maps for the other years analyzed.

@@

      # Heatmap for every combination of year and mode
      assign(paste0("heatmap_plot_", year, "_", mode),
               ggplot(heatmap_melted, aes(x = Destination, y = Origin, fill = value)) +
                  geom_tile(color = "white") +
                  scale_fill_gradientn(colors = RColorBrewer::brewer.pal(9, "YlOrRd"), 
                                       na.value = "gray95", 
                                       limits = c(0, ceiling(max(heatmap_melted$value, na.rm = TRUE))), 
                                       breaks =  c(0, ceiling(max(heatmap_melted$value, na.rm = TRUE))), 
                                       labels = c("0%", paste0(ceiling(max(heatmap_melted$value, na.rm = TRUE)), "%"))) +
                  #geom_text(data = subset(heatmap_melted, !is.na(value) & value >= 0), 
                  #    aes(label = round(value)), 
                  #    size = 4, color = "black") +
                  labs(x = "Destination", y = "Origin", fill = "Percentage", caption = year) +
                  theme_minimal() +
                  theme(
                  axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
                  axis.text.y = element_text(size = 16, margin = margin(l = 15)),
                  axis.title = element_text(size = 16),
                  legend.title = element_text(size = 16, margin = margin(b = 15)), 
                  legend.text = element_text(size = 16),
                  plot.caption = element_text(hjust = 0.5, size = 20, face = "bold", margin = margin(t = 28.35, unit = "pt")))

@@

`ActiveCA` includes a total of
`r nrow(unique(gss_impedances[,c("YEAR","dest_label","f_name","MODE")]))`
pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas. Table \ref{tab:table-04} shows the best impedance functions for walking trips to `Work or school` across survey years, where all distributions, except for a gamma function in 1998, are log-normal.

```{r table-04, echo=FALSE, cache=FALSE, warning=FALSE}
# selecting mode and destination to display impedance functions
mode <- c("Walking","Cycling")
destino <- "Work or school"

# filtering the impedance functions in accordance with the mode/destination setted in the step before and creating a latex table 
selected_functions <- gss_impedances_renamed %>%
  filter(dest_label == destino, MODE %in% mode) %>%
  group_by(MODE, YEAR, dest_label) %>% 
  summarise(Function = first(f_name), 
            est_1 = first(est_1),
            est_2 = first(est_2),
            AIC = first(aic),
            f_name = first(f_name),
            count = first(count),
            .groups = "drop") %>% 
  mutate(Function = case_when(
    Function == "lnorm" ~ "Lognormal",
    Function == "gamma" ~ "Gamma",
    Function == "unif" ~ "Uniform",
    TRUE ~ Function), 
  across(c(est_1,est_2), ~ round(., digits = 2)), 
  across(c(AIC), ~ round(., digits = 0))) %>% 
  rename(Mode = MODE, Year = YEAR, Destination = dest_label) %>%
  unique() %>% 
  arrange(Mode, Year) %>%
  select(Mode, Year, Function, est_1, est_2, AIC, count)

selected_functions_tab <- selected_functions %>% 
  kable(format = "latex",
        booktabs = TRUE,
        caption = paste0("\\label{table-04}Impedance functions and AIC for `", mode ,"` trips considering `", destino,"` as destination."), 
        col.names = c("Mode", "Year", "Impedance function","Parameter 1*","Parameter 2*", "AIC","Count")) %>% 
    kable_styling(full_width = FALSE, 
                latex_options=c("scale_down"),
                font_size = 10) %>% 
  row_spec(0, bold = TRUE, align = "c") %>% 
  collapse_rows(columns = 1, valign = "top", latex_hline = "major") %>%
  footnote(general = c("*For lognormal distributions, 'Parameter 1' and 'Parameter 2' refer to the mean and standard deviation of the distribution on the logarithmic scaler, respectively. For gamma distribution, 'Parameter 1' and 'Parameter 2' refer to the rate and shape of the distribution, respectively. `AIC` means Akaike information criterion."), threeparttable = T)

selected_functions_tab
```

Figure
\ref{fig:figure-03} illustrates these fitted functions (red line)
alongside histograms of empirical travel times (grey bars). As for the
others pre-estimated functions, these examples allow to calculate gravity-based
accessibility measures for active travel across various destinations and temporal scales in Canadian urban areas.

```{r theoretical_impedance_values, include = FALSE}
theoretical_impedance <- function(df_functions, travel_cost) {

  # Initialize the main data frame to store results
  result <- data.frame()

  # Create the 't' sequence
  t_seq <- seq(1, travel_cost, 1)

  for (mode in unique(df_functions$Mode)){

    for (year in unique(df_functions[df_functions$Mode == mode, ]$Year)){

      for (destino in unique(df_functions[df_functions$Year == year & df_functions$Mode == mode, ]$Destination)){

      name_function <- df_functions[df_functions$Year == year & 
                                      df_functions$Mode == mode & 
                                      df_functions$Destination == destino, ]$f_name

      print(paste0("YEAR: ", year, " MODE: ", mode," Destination: ", destino))

      subset <- df_functions %>% 
        filter(Year == year & Mode == mode & Destination == destino)

      # Temporary data frame for this iteration
      temp_df <- data.frame(t = t_seq)

      # Add the appropriate distribution
      if (name_function == "lnorm") {
        temp_df$f <- dlnorm(temp_df$t,
                            meanlog = subset$est_1[1],
                            sdlog = subset$est_2[1])

      } else if (name_function == "gamma") {  
        temp_df$f <- dgamma(temp_df$t,
                            shape = subset$est_1[1],
                            rate = subset$est_2[1])

      } else if (name_function == "norm") {  
        temp_df$f <- dnorm(temp_df$t,
                           mean = subset$est_1[1],
                           sd = subset$est_2[1])

      } else if (name_function == "exp") {  
        temp_df$f <- dexp(temp_df$t,
                          rate = subset$est_1[1])

      } else if (name_function == "unif") {  
        temp_df$f <- dunif(temp_df$t,
                           min = subset$est_1[1],
                           max = subset$est_2[1])
      }

      # Add mode, year and destination columns
      temp_df$MODE <- mode
      temp_df$YEAR <- year
      temp_df$dest_label <- destino

      # Append to the result data frame
      result <- rbind(result, temp_df)
    }
  }
}
  return(result)
}

max_duration <- 100
